---
title: "Environmental covariates, PCA and MESS analysis prior *Anopheles* distribution mapping in Kasaï-Central"
format: pdf
editor: visual
author: "Victoire Nsabatien & all supervisors"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Introduction

This document presents the environmental covariates, principal component analysis (PCA), and multivariate environmental similarity surface (MESS) conducted prior to the preliminary mapping of *Anopheles* distribution in Kasaï-Central.

# Surveyed households

Surveys were conducted across Kasaï-Central Province, Democratic Republic of the Congo. The figure shows the location of surveyed households across the province.

```{r setup, include=FALSE}
library(dplyr)
library(terra)
library(ggplot2)
library(tidyterra)
library(geodata)
library(readr)
library(cowplot)
library(idpalette)
library(mgcv)
```

```{r}
#| fig-cap: "Surveyed household locations in Kasaï-Central shown as pink points."

library(cowplot)

coords <- read_csv("data/clean/kc_household_coords.csv",
                   col_types = cols(
                     health_zone = col_character(),
                     health_area = col_character(),
                     village = col_character(),
                     house_number = col_character(),
                     lat_dd = col_double(),
                     long_dd = col_double()
                   ))

drc <- geodata::gadm("COD", level = 0)
drc_l1 <- geodata::gadm("COD", level = 1)
drc_l2 <- geodata::gadm("COD", level = 2)

kc <- drc_l1 |>
  filter(NAME_1 == "Kasaï-Central")

kc_extent <- ext(kc)

p1 <- ggplot() +
  geom_spatvector(
    data = crop(drc_l2, kc_extent),
    fill = "grey98",
    col = "grey30",
    alpha = 0.9
  ) +
  geom_spatvector(
    data = kc,
    fill = "transparent",
    col = "grey5",
    linewidth = 0.75
  ) +
  geom_point(
    data = coords,
    aes(x = long_dd, y = lat_dd),
    col = "hotpink",
    size = 0.6
  ) +
  theme_void()

p2 <- ggplot() +
  geom_spatvector(
    data = drc_l1,
    fill = "grey98",
    col = "grey30"
  ) +
  geom_spatvector(
    data = drc,
    fill = "transparent",
    col = "grey5",
    linewidth = 0.75
  ) +
  geom_spatvector(
    data = kc,
    fill = "grey98",
    col = "grey5",
    alpha = 0.9
  ) +
  geom_point(
    data = coords,
    aes(x = long_dd, y = lat_dd),
    col = "hotpink",
    size = 0.6
  ) +
  theme_void()

ggdraw(p2) +
  draw_plot(
    p1,
    x = 0.38,
    y = 0,
    width = 0.62,
    height = 0.62
  )
```
